---
title: "Recoding data"
subtitle: "<br><br> Environmental Data Science"
author: "[https://enst222.github.io/website/](https://enst222.github.io/website/)"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

---

### There are a LOT of different ways to recode and transform your data in R. 

We will cover a few of them, but Google is your friend to figure out how to recode in your own unique use cases.
---

## Code-along

.pull-left[
First load packages:
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```
]

.pull-right[
Then tidy the data:
```{r}
iris <- as_tibble(iris) %>%
  clean_names()

iris
```
]
---

### Recode species so that setosa = "a", versicolor = "b", and virginica = "c"

.pull-left[
```{r}
distinct(iris, species)
```
]
---

## Use `mutate()` and `case_match()` to replace or "recode" values

```{r}
iris %>%
  mutate(species = case_match(species, "setosa" ~ "a", #<<
                              "versicolor" ~ "b", #<<
                              "virginica" ~ "c")) %>% #<<
  distinct(species)
```

---

## Use .default argument if you want to keep some values the same

```{r}
iris %>%
  mutate(species = case_match(species, "setosa" ~ "a",
                              .default = species)) %>% #<<
  distinct(species)
```

---

## Use `mutate()` and `if_else()` if you want to replace values based on certain conditions

If `petal_length` is less than three, recode `petal_length` to "short". Otherwise recode `petal_length` to "long".
```{r}
iris
```

---

## Use `mutate()` and `if_else()` if you want to replace values based on certain conditions

If `petal_length` is less than three, recode `petal_length` to "short". Otherwise recode `petal_length` to "long".
```{r}
iris %>%
  mutate(petal_length = if_else(petal_length < 3, "short", "long")) #<<
```

---

## Create a new variable if you want to preserve the original values

```{r}
iris %>%
  mutate(petal_category = if_else(petal_length < 3, "short", "long")) #<<
```

---

## Use `case_when()` for more complicated conditional statements

If `petal_length` is less than or equal to 2, recode `petal_length` to "small". If `petal_length` is greater than 2 and less than or equal to 5, recode to "medium". If `petal_length` is greater than 5, recode to "large".
```{r}
iris %>%
  mutate(petal_category = case_when(petal_length <= 2 ~ "small", #<<
                                    petal_length > 2 & sepal_length <= 5 ~ "medium", #<<
                                    petal_length > 5 ~ "large")) %>% #<<
  distinct(petal_category)
```

---

### Use `TRUE` to index values that were not already defined

For all other values of `petal_length`, recode to "other".
```{r}
iris %>%
  mutate(petal_category = case_when(petal_length <= 2 ~ "small", 
                                    petal_length > 2 & sepal_length <= 5 ~ "medium", 
                                    petal_length > 5 ~ "large",
                                    TRUE ~ "other")) %>% #<<
  distinct(petal_category)
```

---

class:middle

# An example, with some data wrangling too

---

### Create new categorical variable, then group and summarize by that variable

Calculate number of category 3, 4, and 5 storms each decade and plot the results. 
```{r}
storms %>%
  select(name, year, month, day, hour, category)
```

---

## First filter so year is between 1980 and 2020 and category >= 3 

```{r}
storms %>%
  select(name, year, month, day, hour, category) %>% #<<
  filter(year >= 1980, #<<
         year < 2020, #<<
         category > 3) #<<
```

---

### Then group by year and storm name and calculate the max category reached for each storm and year

.panelset[

.panel[.panel-name[Code]
```{r, eval = FALSE}
storms %>%
  select(name, year, month, day, hour, category) %>% 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) %>%
  group_by(name, year) %>% #<<
  summarize(maxcat = max(category)) #<<
```
]

.panel[.panel-name[Output]
```{r, echo = FALSE}
storms %>%
  select(name, year, month, day, hour, category) %>% 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) %>%
  group_by(name, year) %>% #<<
  summarize(maxcat = max(category)) #<<
```
]
]

---

### Next, create a `decade` variable using `mutate()` and `case_when()` based on the `year` variable

.panelset[

.panel[.panel-name[Code]
```{r, eval = FALSE}
storms %>%
  select(name, year, month, day, hour, category) %>%
  filter(year >= 1980, 
         year < 2020, 
         category > 3) %>%
  group_by(name, year) %>%
  summarize(maxcat = max(category)) %>%
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s", #<<
                            year %in% 1990:1999 ~ "1990s", #<<
                            year %in% 2000:2009 ~ "2000s", #<<
                            year %in% 2010:2019 ~ "2010s")) #<<
```
]

.panel[.panel-name[Output]
```{r, echo = FALSE}
storms %>%
  select(name, year, month, day, hour, category) %>%
  filter(year >= 1980, 
         year < 2020, 
         category > 3) %>%
  group_by(name, year) %>%
  summarize(maxcat = max(category)) %>%
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s", #<<
                            year %in% 1990:1999 ~ "1990s", #<<
                            year %in% 2000:2009 ~ "2000s", #<<
                            year %in% 2010:2019 ~ "2010s")) #<<
```
]
]

---

### Finally, group by the new decade variable and count the number of storms in each decade

.panelset[

.panel[.panel-name[Code]
```{r, eval = FALSE}
storms %>%
  select(name, year, month, day, hour, category) %>% 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) %>%
  group_by(name, year) %>%
  summarize(maxcat = max(category)) %>%
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s",
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s",
                            year %in% 2010:2019 ~ "2010s")) %>%
  group_by(decade) %>% #<<
  summarize(n = n()) #<<
```
]

.panel[.panel-name[Output]
```{r, echo = FALSE}
storms %>%
  select(name, year, month, day, hour, category) %>% 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) %>%
  group_by(name, year) %>%
  summarize(maxcat = max(category)) %>%
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s",
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s",
                            year %in% 2010:2019 ~ "2010s")) %>%
  group_by(decade) %>% #<<
  summarize(n = n()) #<<
```
]
]

---

### and then plot!

.panelset[

.panel[.panel-name[Code]
```{r "storm-plot", fig.show="hide"}
storms %>%
  filter(year >= 1980,
         year < 2020,
         category > 3) %>%
  group_by(name, year) %>%
  summarize(maxcat = max(category)) %>%
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s",
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s",
                            year %in% 2010:2019 ~ "2010s")) %>%
  group_by(decade) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = decade, y = n)) + #<<
  geom_bar(stat = "identity") + #<<
  labs(x = "", y = "Number of category 3+ hurricanes") + #<<
  theme_light() #<<
```
]

.panel[.panel-name[Plot]
```{r ref.label="storm-plot", echo = FALSE, out.width="60%"}

```
]
]
